\LaTeXe\ guide for the Journal of Functional Programming

### 2019-04-24 papers/jfp/5-rebuilders.tex
Redo the rebuilders section

\todo{Elaborate (it's very terse right now) with examples}

\todo{Add Shake's step traces: Optimisation, Correctness}

\todo{Explain how to "add constructiveness" to a trace system.  Mabye even concretely:  `constructive :: TraceSystem -> TraceSystem`.Maybe collapse verifying traces and constrctive traces, at least in table 2.  (This would have impact on the structore of the whole trace section.)}

@@

One potential implementation would be to record all arguments passed to
\hs{recordVT} in a list, and verify by simply checking if any list item matches
the information passed by \hs{verifyVT}.  We discuss smarter implementations
in~\S\ref{sec-smart-traces}.

In the example above, we have used abstract types for the traces. Concretely,
in our example code, they are all recorded as lists of:

@@

Here \hs{r} is either \hs{Hash}~\hs{v} (verifying traces) or
\hs{v} (constructive traces). A real system is highly likely to use a more
optimised implementation. Some of the most obvious optimisations are:

\begin{itemize}
\item Any system using verifying traces is unlikely to see significant benefit
from storing more than one \hs{Trace} per key\footnote{There is a small chance
of a benefit if the dependencies change but the result does not, and then the
dependencies change back to what they were before.}. Therefore, such systems can
store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v))}, where the
initial \hs{k} is the \hs{key} field of \hs{Trace}.

\item Any system using \hs{Applicative} dependencies can omit the dependency
keys from the \hs{Trace} since they can be recovered from the \hs{key} field.

\item Any \hs{Applicative} build system using constructive traces, e.g.
\CloudBuild, can index directly from the key and results to the output result~--~i.e.
\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces
are stored on a central server, the client can compute the key and the hashes of
its dependencies, then make a single call to the server to retrieve the result.

\item Many cloud build systems store hashes of values in the trace information,
then have a separate content-addressable cache which associates hashes with
their actual contents.
\vspace{-1mm}
\end{itemize}

@@

The \hs{built} field is when the \hs{key} last rebuilt. The \hs{changed} field is when the \hs{result} last changed -- if the last build changed the value, it will be equal to \hs{built}, otherwise it will be older. The \hs{recordVT} function needs to consult the previous \hs{Trace} to know whether to use \hs{built} or the previous \hs{changed} value as the new \hs{changed} value. The \hs{verifyVT} needs to verify that the \hs{built} time of \hs{key} is newer than the \hs{changed} values of all its \hs{depends}. This scheme preserves minimality and early cutoff. A variant with only one \hs{Time} field would loose early cutoff. Furthermore, the \hs{Time} stamp only needs record which execution of the build is running, so every key built in the same run can share the same \hs{Time} value -- it just needs to be monotically increasing between runs.

@@

\subsubsection{Constructive Traces}\label{sec-constructive-traces}

A verifying trace deliberately records only small hashes, so that it can be small.
A \emph{constructive} trace additionally stores the resulting value.
Once we are storing the complete result it makes sense
to record many constructive traces per key, and to share them with other users,
providing cloud-build functionality. We can represent this additional
information by providing the operations:

@@

\subsubsection{Deep Constructive Traces}\label{sec-deep-constructive-traces}

@@

Current build systems using deep constructive traces always record hashes of
terminal \emph{input keys}, but the technique works equally well if we skip any
number of dependency levels (say $n$ levels). The input-only approach is the
special case of $n = \infty$, and constructive traces are the special case of
$n = 1$. When $n \ge 2$, deep constructive traces require the tasks to be
\emph{deterministic}, as otherwise it is possible to violate correctness, as
illustrated by an example in~\S\ref{sec-cloud-aspects}
(see Fig.~\ref{fig-frankenbuild}).

A downside of deep constructive traces is that they cannot support early cutoff
(\S\ref{sec-background-shake}), other than at $n$ levels of dependencies. On the
other hand, these traces are particularly useful for \emph{shallow builds}, as
discussed in~\S\ref{sec-cloud-aspects}.

\subsection{Efficient Data Structures for Traces}\label{sec-smart-traces}

In the examples above, we have used abstract types for the traces. Concretely,
in our example code, they are all recorded as lists of:

\begin{minted}[xleftmargin=10pt]{haskell}
data Trace k v r = Trace { key :: k, depends :: [(k, Hash v)], result :: r }
\end{minted}

Here \hs{r} is either \hs{Hash}~\hs{v} (verifying traces) or
\hs{v} (constructive traces). A real system is highly likely to use a more
optimised implementation. Some of the most obvious optimisations are:

\begin{itemize}
\item Any system using verifying traces is unlikely to see significant benefit
from storing more than one \hs{Trace} per key\footnote{There is a small chance
of a benefit if the dependencies change but the result does not, and then the
dependencies change back to what they were before.}. Therefore, such systems can
store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v))}, where the
initial \hs{k} is the \hs{key} field of \hs{Trace}.

\item Any system using \hs{Applicative} dependencies can omit the dependency
keys from the \hs{Trace} since they can be recovered from the \hs{key} field.

\item Any \hs{Applicative} build system using constructive traces, e.g.
\CloudBuild, can index directly from the key and results to the output result~--~i.e.
\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces
are stored on a central server, the client can compute the key and the hashes of
its dependencies, then make a single call to the server to retrieve the result.

\item Many cloud build systems store hashes of values in the trace information,
then have a separate content-addressable cache which associates hashes with
their actual contents.
\vspace{-1mm}
\end{itemize}

>>>>

@@

One potential implementation would be to record all arguments passed to
\hs{recordVT} in a list, and verify by simply checking if any list item matches
the information passed by \hs{verifyVT}. Concretely,
in our implementations from \S\ref{sec-implementations}, traces are recorded as lists of:

@@

\noindent Where \hs{r} is \hs{Hash}~\hs{v} for verifying traces (and \hs{v} for constructive traces, discussed later in \S\ref{sec-constructive-traces}). A real system is highly likely to use a more optimised implementation.

The first optimisation is that any system using \hs{Applicative} dependencies can omit the dependency
keys from the \hs{Trace} since they can be recovered from the \hs{key} field.

The next optimisation is that there is only very minor benefit from storing more than one \hs{Trace} per key. Therefore, veryifying traces can be stored as \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v))}, where the
initial \hs{k} is the \hs{key} field of \hs{Trace} -- making \hs{verifyVT} fast.
Note that storing only one \hs{Trace} per key means that if the dependencies change but the result does not, and then the
dependencies change back to what they were before, there will not be a valid \hs{Trace} available and the system will have to rebuild, where a complete list of all historical traces would not have had to. However, it also means that the number of \hs{Trace} structures is bounded by the number of distinct keys, regardless of how many builds are executed -- a useful property.

@@

The \hs{built} field is when the \hs{key} last rebuilt. The \hs{changed} field is when the \hs{result} last changed -- if the last build changed the value, it will be equal to \hs{built}, otherwise it will be older. The \hs{recordVT} function needs to consult the previous \hs{Trace} to know whether to use \hs{built} or the previous \hs{changed} value as the new \hs{changed} value. The \hs{verifyVT} needs to verify that the \hs{built} time of \hs{key} is newer than the \hs{changed} values of all its \hs{depends}. This scheme preserves minimality and early cutoff. A variant with only one \hs{Time} field would loose early cutoff, and indeed corresponds quite closely to \Make. Furthermore, the \hs{Time} stamp only needs record which execution of the build is running, so every key built in the same run can share the same \hs{Time} value -- it just needs to be monotically increasing between runs.

@@

\subsection{Constructive Traces}\label{sec-constructive-traces}

A verifying trace deliberately records only small hashes, so that it can be small.
In contrast, a \emph{constructive} trace also stores the resulting value. Concretely,
it records a list of \hs{Trace}~\hs{k}~\hs{v}.
Once we are storing the complete result it makes sense
to record many constructive traces per key, and to share them with other users,
providing cloud-build functionality. We can represent this additional
information with the operations:

@@

Any \hs{Applicative} build system using constructive traces, e.g.
\CloudBuild, can index directly from the key and results to the output result~--~i.e.
\hs{Map}~\hs{(@@k,}~\hs{[Hash}~\hs{v])}~\hs{v}. Importantly, assuming the traces
are stored on a central server, the client can compute the key and the hashes of
its dependencies, then make a single call to the server to retrieve the result.

In practice, many cloud build systems store hashes of values in the trace information,
then have a separate content-addressable cache which associates hashes with
their actual contents.

\subsection{Deep Constructive Traces}\label{sec-deep-constructive-traces}

@@

Concretely, taking the example from Figure \ref{fig-make}, to decide whether \hs{main.exe}
is out of data, a \emph{constructive} trace would look at \hs{util.o} and \hs{main.o}
(the immediate dependencies), while a \hs{deep constructive} trace would look at \hs{util.c},
\hs{util.h} and \hs{main.c}.

An advantage of deep constructive traces is that to decide if \hs{main.exe} is up to date only
requires consulting its inputs, not even considering \hs{util.o} or \hs{main.o}.
Such a feature is often known as a \emph{shallow build}, as discussed in~\S\ref{sec-cloud-aspects}.

There are two primary disadvantages of deep constructive traces:

\begin{description}
\item[Deterministic tasks] If the tasks are not \emph{deterministic} then it is possible to
violate correctness, as illustrated by an example in~\S\ref{sec-cloud-aspects}
(see Fig.~\ref{fig-frankenbuild}).
\item[No early cutoff] Such traces cannot support early cutoff
(\S\ref{sec-background-shake}), since the results of intermediate computations are not considered.
\end{description}

Current build systems using deep constructive traces always record hashes of
terminal \emph{input keys}, but the technique works equally well if we skip any
number of dependency levels (say $n$ levels). The input-only approach is the
special case of $n = \infty$, and constructive traces are the special case of
$n = 1$. By picking values of $n$ in between we would regain some early cutoff, at the
cost of losing such simple shallow builds and still requiring determinstic tasks.
